Accessible, realistic genome simulation with selection using \stdpopsim

### 2025-03-17 analysis2_manuscript.tex
updated my affiliation and added some explanation to ROC curves

% author tranche explanation:
% co-first authors
\author[1,*]{Graham Gower}
\author[2,*]{Nathaniel S. Pope}
\author[2,*]{Murillo F. Rodrigues}
\author[2,*]{Silas Tittes}
\author[3,*]{Linh N. Tran}

@@

\affil[*]{\small{These authors contributed equally to the paper.}}
\affil[1]{\small{Section for Molecular Ecology and Evolution, Globe Institute, University of Copenhagen, Denmark}}
\affil[2]{\small{Institute of Ecology and Evolution, University of Oregon, Eugene OR 97402, USA}}
\affil[3]{\small{Department of Molecular and Cellular Biology, University of Arizona, Tucson, AZ 85721, USA}}
\affil[4]{\small{Center for Systems Biology, New York University, New York NY, 10012 USA}}
\affil[5]{\small{Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles CA, USA}}
\affil[6]{\small{The Jackson Laboratory, Bar Harbor, ME, 04609 USA}}
\affil[7]{\small{Department of Computational Biology, Cornell University, Ithaca NY, USA}}
\affil[8]{\small{Department of Evolutionary Anthropology, University of Vienna, Vienna, Austria}}
\affil[9]{\small{Human Evolution and Archaeological Sciences (HEAS), University of Vienna, Vienna, Austria}}
\affil[10]{\small{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford OX3 7LF, UK}}
\affil[11]{\small{Instituto Nacional de Investigación Agropecuaria (INIA), Estación Experimental Las Brujas, Ruta 48 km 10, Canelones, Uruguay}}
\affil[12]{\small{Department of Genome Sciences, University of Washington, Seattle WA, USA}}
\affil[13]{\small{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, Edinburgh EH25 9RG, UK}}
\affil[14]{\small{Department of Biology, Pennsylvania State University, University Park, PA 16802, USA}}
\affil[15]{\small{Department of Integrative Biology, University of Wisconsin-Madison, Madison WI, USA}}
\affil[16]{\small{Efi Arazi School of Computer Science, Reichman University, Herzliya, Israel}}
\affil[17]{\small{Department of Genetics, University of North Carolina at Chapel Hill, Chapel Hill NC 27599, USA}}
\affil[18]{\small{Department of Data Science, University of Oregon, Eugene OR 97402, USA}}

@@

    We compared the overall performance of the three sweep-detection methods using
    receiver operating characteristic (ROC) curves (Figure \ref{fig:roc-curves}).
    Performance varies for each of the sampled populations.
    In the CEU sample, \sweepfinder and \diploshic perform similarly with a high area under the curve (AUC) score,
    (AUC = 0.906 and 0.903 respectively), but both underperform the simple method
    that uses reduced diversity ($\pi$) as an indicator (AUC = 0.932).
    In the YRI sample, \diploshic outperforms the other two methods,
    with an AUC of 0.979, followed by using outliers of $\pi$ with an AUC of 0.957,
    and then \sweepfinder with an AUC of 0.939.
    We note that we trained \diploshic on simulations under a constant size population,
    which more closely matches the demographic history of YRI, than that of CEU, which 
    is characterized by a strong recent bottleneck followed by population expansion.
    %\drscomment{note: I am couching the language above a bit because we have previously shown that shic doesn't suffer in the sweep vs. nonsweep question when doing other train-on-constant-test-on-nonequilib benchmarks. Could be something else going on here (or some interaction between demog misspec and rec rate variation), but in any case I am not certain of the cause so I am toning down our certainty slightly.}
    Interestingly, the simple method that uses reduced diversity ($\pi$) as an indicator
    for a selective sweep performs very well overall (especially on the CEU sample).
    However, its accuracy is much more variable along the chromosome compared to the other two method (see Figure \ref{fig:chr1_power}),
    and for quite a few locations along the chromosome it has power below $50\%$
    when applied to both sample sets.
    \sweepfinder appears to have the least variation in power along the chromosome or across populations.

@@

% roc curves for sweeps
\begin{figure}
    \centering
    \includegraphics[width=0.8 \textwidth]{figures/sweeps/roc_neutral_null.pdf}
    \caption{
    \label{fig:roc-curves}
    Receiver operating characteristic (ROC) curves for the three methods for detecting sweeps.
    True positive rates (TPR) were computed across all 100,000 datasets simulated with selective sweeps,
    and false positive rates (FPR) were established by applying the methods to genomes simulated
    under neutral evolution with the same demographic history and genetic map
    used in the simulations that included sweeps (see \textbf{Methods}).
    AUC scores are shown for each method. \sweepfinder is labelled CLR in the figure.
    }
    %DRS-fig: can we include AUCs here?
\end{figure}

>>>>

% author tranche explanation:
% co-first authors
\author[1,*]{Graham Gower}
\author[2,*]{Nathaniel S. Pope}
\author[2,19,*]{Murillo F. Rodrigues}
\author[2,*]{Silas Tittes}
\author[3,*]{Linh N. Tran}

@@

\affil[*]{\small{These authors contributed equally to the paper.}}
\affil[1]{\small{Section for Molecular Ecology and Evolution, Globe Institute, University of Copenhagen, Denmark}}
\affil[2]{\small{Institute of Ecology and Evolution, University of Oregon, Eugene OR 97402, USA}}
\affil[3]{\small{Department of Molecular and Cellular Biology, University of Arizona, Tucson, AZ 85721, USA}}
\affil[4]{\small{Center for Systems Biology, New York University, New York NY, 10012 USA}}
\affil[5]{\small{Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles CA, USA}}
\affil[6]{\small{The Jackson Laboratory, Bar Harbor, ME, 04609 USA}}
\affil[7]{\small{Department of Computational Biology, Cornell University, Ithaca NY, USA}}
\affil[8]{\small{Department of Evolutionary Anthropology, University of Vienna, Vienna, Austria}}
\affil[9]{\small{Human Evolution and Archaeological Sciences (HEAS), University of Vienna, Vienna, Austria}}
\affil[10]{\small{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford OX3 7LF, UK}}
\affil[11]{\small{Instituto Nacional de Investigación Agropecuaria (INIA), Estación Experimental Las Brujas, Ruta 48 km 10, Canelones, Uruguay}}
\affil[12]{\small{Department of Genome Sciences, University of Washington, Seattle WA, USA}}
\affil[13]{\small{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, Edinburgh EH25 9RG, UK}}
\affil[14]{\small{Department of Biology, Pennsylvania State University, University Park, PA 16802, USA}}
\affil[15]{\small{Department of Integrative Biology, University of Wisconsin-Madison, Madison WI, USA}}
\affil[16]{\small{Efi Arazi School of Computer Science, Reichman University, Herzliya, Israel}}
\affil[17]{\small{Department of Genetics, University of North Carolina at Chapel Hill, Chapel Hill NC 27599, USA}}
\affil[18]{\small{Department of Data Science, University of Oregon, Eugene OR 97402, USA}}
\affil[19]{\small{Division of Genetics, Oregon National Primate Center, Oregon Health and Science University, Beaverton OR 97006, USA}}

@@

    We compared the overall performance of the three sweep-detection methods using
    receiver operating characteristic (ROC) curves (Figure \ref{fig:roc-curves}).
    Performance varies for each of the sampled populations.
    In the CEU sample, \sweepfinder and \diploshic perform similarly with a high area under the curve (AUC) score,
    (AUC = 0.906 and 0.903 respectively), but both underperform the simple method
    that uses reduced diversity ($\pi$) as an indicator (AUC = 0.932).
    In the YRI sample, \diploshic outperforms the other two methods,
    with an AUC of 0.979, followed by using outliers of $\pi$ with an AUC of 0.957,
    and then \sweepfinder with an AUC of 0.939.
    Note, however, that despite the apparent good performance of the $\pi$ statistic,
    it showed extreme heterogeneity across the genome. % MFR: I think this is an important point. Basically there is a Simpson's paradox here.
    We note that we trained \diploshic on simulations under a constant size population,
    which more closely matches the demographic history of YRI, than that of CEU, which 
    is characterized by a strong recent bottleneck followed by population expansion.
    %\drscomment{note: I am couching the language above a bit because we have previously shown that shic doesn't suffer in the sweep vs. nonsweep question when doing other train-on-constant-test-on-nonequilib benchmarks. Could be something else going on here (or some interaction between demog misspec and rec rate variation), but in any case I am not certain of the cause so I am toning down our certainty slightly.}
    Interestingly, the simple method that uses reduced diversity ($\pi$) as an indicator
    for a selective sweep performs very well overall (especially on the CEU sample).
    However, its accuracy is much more variable along the chromosome compared to the other two method (see Figure \ref{fig:chr1_power}),
    and for quite a few locations along the chromosome it has power below $50\%$
    when applied to both sample sets.
    \sweepfinder appears to have the least variation in power along the chromosome or across populations.

@@

% roc curves for sweeps
\begin{figure}
    \centering
    \includegraphics[width=0.8 \textwidth]{figures/sweeps/roc_neutral_null.pdf}
    \caption{
    \label{fig:roc-curves}
    Receiver operating characteristic (ROC) curves for the three methods for detecting sweeps.
    True positive rates (TPR) were computed across all 100,000 datasets simulated with selective sweeps,
    and false positive rates (FPR) were established by applying the methods to genomes simulated
    under BGS applied on exons with the same demographic history and genetic map
    used in the simulations that included sweeps (see \textbf{Methods}).
    AUC scores are shown for each method. \sweepfinder is labelled CLR in the figure.
    }
    %DRS-fig: can we include AUCs here?
\end{figure}
